---
title: "TITLETK"
author: "Jonah Wolf"
date: "5/28/2020"
output:
  pdf_document:
    latex_engine:
      xelatex
bibliography: Final-Outline.bib
nocite: '@*'
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
setwd("~/Documents/School/John Jay/Research Methods/Research-Methods-I")
library(DBI)
library(tidyverse)
library(reshape2)
library(stargazer)
library(future)
```

```{r, echo=FALSE, message=FALSE}

# Read Pitchfork dataset
p4k_future <- future({
  con <- dbConnect(RSQLite::SQLite(), "./Datasets/pitchfork-database.sqlite")
  pitchfork <- dbReadTable(con, "reviews")
  genres <- dbReadTable(con, "genres")
  dbDisconnect(con)

  # Tidy Pitchfork dataset
  pitchfork <- pitchfork %>%
    unique() %>%
    mutate(pub_date = as.Date(pub_date))
  
  genres <- genres %>%
    unique %>%
    pivot_wider(id_cols = reviewid, names_from = genre, values_from = genre) %>%
    select(-matches("NA")) %>%
    rename(
      pop = `pop/r&b`,
      country = `folk/country`
    )
  
  genres[2:10] <- lapply(genres[2:10], function(x) (x != "NULL"))
  
  merge(pitchfork, genres, by = "reviewid")
})

pitchfork <- value(p4k_future)
  
# Read Billboard dataset
bb_future <- future({
  con2 <- dbConnect(RSQLite::SQLite(), "./Datasets/billboard-200.db")
  billboard <- dbReadTable(con2, "albums")
  dbDisconnect(con2)
  return(billboard)
})  
  
# Tidy Billboard dataset
billboard <- value(bb_future) %>%
  slice(2:n()) %>%
  rename(title = album) %>%
  mutate(
    title = tolower(title),
    artist = tolower(artist),
    rank = as.integer(rank),
    date = as.Date(date, format = "%Y-%m-%d")
    ) %>%
  group_by(title, artist) %>%
  summarize(
    rank.mean = mean(rank),
    rank.peak = min(rank),
    date.entered = min(date),
    date.left = max(date),
    weeks = n()
    )

# Merge datasets
pitchfork <- left_join(pitchfork, billboard, by = c("title", "artist")) %>%
  mutate(
    charted = ! is.na(date.entered),
    reviewed_first = date.entered > pub_date
    )
```

# Introduction

This paper attempts to analyze the relationship between elite and popular musical tastes by comparing sales data from the Billboard 200 album chart with album scores from the reviews site Pitchfork for the years 1999--2017. While empirical research such as [@Dewan] and [@Robette] has attempted to quantify the effect of media on music popularity, this paper inverts the relationship to ask how critics respond to mass tastes. In so doing, it hopes to add empirical knowledge to the debates around "poptimism" and "rockism" that permeated metacritical discourse in the early 21st century (see notably [@Sanneh]).

Pitchfork describes itself to advertisers as "the most trusted voice in music" [@Pitchfork]. Nonetheless, its influence is difficult to quantify. As [@Robette] writes, "criticism is part of an ensemble of practices of intermediation" (p. 121). He and [@Beck] cite the same two examples---a 9.7 score for the Arcade Fire [@Moore] and an 0.0 for Travis Morrison [@Dahlen]---for how the site's reviews could affect each artists' careers.

While [@Robette] describes how Pitchfork transmitted its influence through other sections of the music industry such as record stores, labels, and other magazines (pp. 123--4), and produces a multiple correspondence analysis showing an increasing overlap between albums cited in Pitchfork's year-end lists and those of other publications (pp. 133--5).  [@Beck] notes how the decline of these institutions, amid the rise of digital music platforms, only amplified Pitchfork's influence: "In the 21st century, we are all record store clerks" (p. 179).

There is a paradox, however, at the root of this influence. As [@Frith] puts it, "it's the writers' independence from record company influence that gives good reviewers their authenticity" (p. 174). The loyalty commanded by any critic depends equally on her independence from prevailing tastes, and her reliability in satisfying the tastes of her readership. This paradox is only amplified in the field of rock music, and even more so in indie rock---forms that embody "a tension between the music's aspirations to art and its reality of being a mass, commercial culture" [@Hamilton, p. 80]. Moreover, "underground papers were important as the source of what become the dominant ideology of rock" ([@Frith] p. 169).

# Theoretical framework

Frith's description of the professionalization of *Rolling Stone* magazine in the 1970s, as it "became integrated into the American music business" (p. 170) anticipates Pitchfork's trajectory in the 2000s. [@Robette] notes that, between 2003 and 2013, the share of Billboard-charting albums on Pitchfork's year-end lists increased from 17 to 30%. He proposes "a movement of convergence between Pitchfork and the modal taste" (p. 131).

Describing this phenomenon as a "convergence," Robette avoids one-sided explanations of accomodation on the part of either Pitchfork or its audience. Whatever influence the publication had was matched by a broadening of the music it covered. The share of Pitchfork reviews in the "pop/R&B" genre grew at an average annual rate of 8.9% in the years 1999 through 2016, while the share in the "rock" genre declined by an average 3.2%.

This shift can be described by ideological as much as by mercenary reasons. In the period described by this study, numerous critics identified a shift in elite musical ideology---what [@Petersen and Kern] call "from snob to omnivore" (1996). Petersen and Kern suggest five possible causes for this shift: "structural change" (whereby geographic migration and mass media expose listeners to other cultures), "value change" (the abandonment of status hierarchies of gender, race, religion, and ethnicity), "art-world change" (a shift from conservatizing to "avant-guardist" [sic] ideologies), and "generational politics" (the baby boomers' continued attachment to the culture of their youth) (p. 905).

Beck drolly charts the broadening of Pitchfork's coverage beyond white suburban indie rock to include rap, jazz ("'Trane takes it to heaven and back with some style, man" [@Schreiber]), and dance music ("You people at shows who don't dance, who don't know a good time, who can't have fun, who sneer and scoff at the supposed inferior---it's you this music strikes a blow against" [@Schreiber 2003]).

# Data

This paper uses a dataset of 18,389 album reviews published on Pitchfork from 1999 through 2016 ([@Conaway 2017]). The publication and its authors rate albums on a decimal scale, from 0.0 to 10.0; chosen releases are also assigned the tag "Best New Music." Albums are tagged with the following genre labels: "electronic\", \"metal\", \"rock\", \"rap\", \"experimental\", \"pop/r&b\", \"folk/country\", \"jazz\", and \"global."

The mean overall score is 7.006, and the median 7.2. For "Best New Music" releases, the mean is 8.676 and median 8.6.^[These figures are depressed somewhat, as the "Best New Music" label is sometimes applied across a group of differently scored albums by the same artist. Weezer's 2010 album _Death to False Metal_, for example, was reviewed alongside a reissue of their classic _Pinkerton_, and so received a "Best New Music" tag despite its score of 3.5 ([@Cohen 2010]).]

```{r}
paste(colnames(genres[-1]), collapse = '", "')

summary(pitchfork$score)

ggplot(data = pitchfork, aes(x = score)) +
  geom_histogram(binwidth = .5)

bnm <- pitchfork %>% filter(best_new_music == TRUE)

summary(bnm$score)

ggplot(data = bnm, aes(x = score)) +
  geom_histogram(binwidth = .5)
```

"Rock" has consistently dominated the distribution of genres, though its share of total reviews fell from a majority of 65.2% in 2000 to a strong plurality of 34.2% in 2016.

## graph share of each genre by year

```{r}
pf_years <- pitchfork %>%
  group_by(pub_year) %>%
  filter(pub_year < 2017) %>%
  summarize(
    total = n(),
    across(electronic:global, ~ sum(.x, na.rm = TRUE)),
    .groups = "drop"
    ) %>%
  mutate(
    across(electronic:global, ~ .x / total)
    )

summary(pf_years$rock)

shares <- pf_years %>%
  select(-total) %>%
  melt(id.vars = "pub_year")

ggplot(data = shares, mapping = aes(x = pub_year)) +
  geom_line(aes(y = value, color = variable)) +
  labs(x = "year", y = "share of total reviews published")
```


```{r}

pf_years_growth <- pf_years %>%
  mutate(
    across(electronic:global, ~ (.x - lag(.x))/lag(.x))
  )

growth_means <- pf_years_growth %>% 
  select(c(3:11)) %>%
  colMeans(na.rm = TRUE)

## make bar chart of growth rates

stargazer(growth_means)
```

```{r}
# box plot scores
```


* [Acoustic and meta features of albums and songs on the Billboard 200 (1963--2019)](https://components.one/datasets/billboard-200/)

```{r}

pf_charted <- pitchfork %>%
  group_by(pub_year) %>%
  filter(pub_year < 2017) %>%
  summarize(
    total = n(),
    charted = sum(charted, na.rm = TRUE),
    .groups = "drop"
    ) %>%
  mutate(
    charted = charted / total
    ) %>%
  select(-total)

# data to add: median scores, chart positions, weeks charted

    # pop_charted = sum(charted[pitchfork$pop], na.rm = TRUE),
    # bnm = sum(best_new_music, na.rm = TRUE),
    # bnm_pop = sum(best_new_music[pitchfork$pop], na.rm = TRUE),
    # bnm_charted = sum(best_new_music[pitchfork$charted], na.rm = TRUE),
    # med_score = median(score, na.rm = TRUE),
    # med_pop = median(score[pitchfork$pop], na.rm = TRUE),
    # med_charted = median(score[pitchfork$charted], na.rm = TRUE)

  #   charted_share = charted / total,
  #   chart_growth = (charted_share - lag(charted_share))/lag(charted_share),
  #   bnm_pop_share = bnm_pop / bnm,
  #   bnm_charted_share = bnm_charted / bnm

# make chart


# meds_long <- pf_years %>%
#   select(pub_year, med_pop, med_charted, med_score) %>%
#   melt(id.vars = "pub_year", data.vars = c("med_pop", "med_charted", "med_score"))
# 
# ggplot(data = meds_long, mapping = aes(x = pub_year)) +
#   geom_line(aes(y = value, linetype = variable)) +
#   labs(x = "year", y = "median score")

```

# Econometric Models and Estimation Methods

$$
\begin{aligned}
  score = f(
    &weeks * electronic * metal * rock * rap * experimental * pop * country + \\
    &reviewed.first * electronic * metal * rock * rap * experimental * pop * country + \\
    &rank.peak * electronic * metal * rock * rap * experimental * pop * country + \\
    &electronic + metal + rock + rap + experimental + pop + country)
\end{aligned}
$$

```{r}
# lm(score ~ weeks * electronic * metal * rock * rap * experimental * pop * country * jazz * global * reviewed_first, merged)

# account for different efforts by artist?
```



# Results

# Conclusions

# Bibliography